---
title: "Co-occurrence Networks with IMDB Movie Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Co-occurrence Networks with IMDB Movie Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 6,
  message = FALSE, warning = FALSE
)
```

This tutorial demonstrates every feature of the `cooccur` package using 1,000 highly-rated IMDB movies (rating $\geq$ 7.0, $\geq$ 1,000 votes, 1970--2024). We build genre co-occurrence networks, actor collaboration networks, compare decades, apply different similarity measures and counting methods, and export to Gephi, igraph, and cograph.

## Data

The dataset includes two tables:

- **movies**: 1,000 films with title, year, genres (comma-delimited), decade, rating, and vote count.
- **actor_movie**: ~25,000 actor--movie pairs linking actors to the films they are "known for" on IMDB.

```{r load-data}
library(cooccur)

movies <- read.csv(system.file("extdata/imdb_movies.csv", package = "cooccur"))
actor_movie <- read.csv(system.file("extdata/imdb_actor_movie.csv", package = "cooccur"))

cat(sprintf("Movies: %d | Actor-movie pairs: %d | Unique actors: %d\n",
            nrow(movies), nrow(actor_movie), length(unique(actor_movie$actor))))
head(movies[, c("primaryTitle", "startYear", "genres", "averageRating")])
```

## 1. Genre co-occurrence (delimited field)

The simplest use case. Each movie has a comma-delimited `genres` column. Genres that appear in the same movie are connected.

```{r genre-basic}
genre_net <- cooccurrence(movies, field = "genres", sep = ",")
genre_net
```

Drama dominates because it's the most frequent genre. Let's normalize with Jaccard similarity to control for popularity:

```{r genre-jaccard}
genre_jac <- co(movies, field = "genres", sep = ",", similarity = "jaccard")
genre_jac
```

Now the top edges reflect genuine affinity rather than just frequency. War--History films co-occur more often than chance would predict given their individual frequencies.

### Comparing similarity measures

Each similarity measure answers a different question about the relationship between two items. Here we compute the top 3 genre pairs under each measure:

```{r similarity-comparison}
sims <- c("none", "jaccard", "cosine", "inclusion",
           "association", "dice", "equivalence")
comparison <- do.call(rbind, lapply(sims, function(s) {
  res <- co(movies, field = "genres", sep = ",", similarity = s, top_n = 3)
  data.frame(similarity = s, from = res$from, to = res$to,
             weight = round(res$weight, 4), count = res$count,
             stringsAsFactors = FALSE)
}))
comparison
```

Notice how different measures surface different structures:

- **`none`** (raw counts) is dominated by Drama because it appears in the most movies --- high frequency drives high co-occurrence regardless of affinity.
- **`jaccard`** controls for frequency by asking: "of all movies that have genre A *or* genre B, how many have *both*?"
- **`inclusion`** gives 1.0 when one genre almost always appears with another --- it reveals subset relationships (e.g., Film-Noir is almost always tagged as Crime or Drama).
- **`association`** penalizes high-frequency genres the most aggressively, surfacing niche pairings that co-occur more than expected under independence.

### Which similarity to use?

The choice depends on what you're looking for:

- **Exploratory work**: Start with `"none"` to see raw counts and understand the data, then try `"jaccard"` or `"cosine"` for a balanced view.
- **Bibliometric and scientometric networks**: `"association"` is recommended by van Eck & Waltman (2009) because it correctly accounts for the expected number of co-occurrences under independence. Two items that are both very frequent will naturally co-occur often; association strength discounts this, revealing which pairs co-occur *more than chance alone would predict*.
- **Detecting hierarchical or subset structure**: `"inclusion"` (the Simpson coefficient) reveals when one item almost always appears with another --- useful for identifying items that are subsets of broader categories, or for detecting dependency relationships.
- **Binary presence/absence networks**: `"jaccard"` or `"dice"` when you only care *whether* items co-occur, not *how often*. Jaccard is stricter (penalizes unbalanced pairs more); Dice is more lenient.
- **Scale-invariant comparison**: `"cosine"` (Salton's cosine) is invariant to the absolute frequency of items --- useful when comparing co-occurrence patterns across datasets of different sizes.
- **Strict filtering**: `"equivalence"` (cosine squared) amplifies differences --- pairs with weak overlap get pushed closer to zero, making it useful when you want to retain only the strongest associations.

## 2. Counting methods

By default, each movie contributes equally to co-occurrence: if a movie has genres A, B, and C, each pair (A-B, A-C, B-C) adds 1. But movies with many genres inflate the network --- a movie tagged with 5 genres creates 10 pairs, while a movie with 2 genres creates just 1.

**Fractional counting** addresses this by weighting each pair by $1/(n-1)$, where $n$ is the number of items in the transaction. A 5-genre movie contributes $1/4 = 0.25$ per pair instead of 1.

```{r counting-full}
# Full counting (default): each pair adds 1
genre_full <- co(movies, field = "genres", sep = ",", top_n = 5)
genre_full
```

```{r counting-fractional}
# Fractional counting: each pair adds 1/(n-1)
genre_frac <- co(movies, field = "genres", sep = ",",
                 counting = "fractional", top_n = 5)
genre_frac
```

Compare how the rankings change:

```{r counting-compare}
full <- co(movies, field = "genres", sep = ",", similarity = "jaccard", top_n = 10)
frac <- co(movies, field = "genres", sep = ",", similarity = "jaccard",
           counting = "fractional", top_n = 10)
merge(
  data.frame(pair = paste(full$from, full$to, sep = "--"),
             weight_full = round(full$weight, 4)),
  data.frame(pair = paste(frac$from, frac$to, sep = "--"),
             weight_frac = round(frac$weight, 4)),
  by = "pair", all = TRUE
)
```

Fractional counting is particularly important in bibliometric networks where some documents have dozens of keywords while others have just two. Without it, keyword-rich documents dominate the network structure.

## 3. Scaling

Raw counts or similarity values can span a wide range. Scaling compresses the distribution for visualization or downstream analysis.

```{r scaling}
# Log scaling: compresses heavy-tailed distributions
genre_log <- co(movies, field = "genres", sep = ",", scale = "log")
head(genre_log)

# Min-max to [0, 1]: makes weights comparable across networks
genre_mm <- co(movies, field = "genres", sep = ",",
               similarity = "jaccard", scale = "minmax")
head(genre_mm)

# Binary: presence/absence of co-occurrence (useful for graph density analysis)
genre_bin <- co(movies, field = "genres", sep = ",", scale = "binary")
head(genre_bin)

# Square root: mild compression, preserves more variation than log
genre_sqrt <- co(movies, field = "genres", sep = ",", scale = "sqrt")
head(genre_sqrt)
```

Scaling can be combined with any similarity measure:

```{r scaling-combined}
# Association strength, log-scaled, min 20 occurrences
co(movies, field = "genres", sep = ",",
   similarity = "association", scale = "log", min_occur = 20, top_n = 5)
```

## 4. Filtering

Three filtering mechanisms control which edges appear in the result.

### Minimum frequency

Drop genres appearing in fewer than 20 movies. This removes rare genres that produce noisy co-occurrence patterns:

```{r min-occur}
genre_filtered <- co(movies, field = "genres", sep = ",",
                     similarity = "jaccard", min_occur = 20)
genre_filtered
```

### Threshold

Keep only edges with Jaccard similarity above 0.15. Applied *after* similarity normalization and scaling:

```{r threshold}
genre_strong <- co(movies, field = "genres", sep = ",",
                   similarity = "jaccard", threshold = 0.15)
genre_strong
```

### Top N edges

Keep the 10 strongest connections:

```{r top-n}
genre_top10 <- co(movies, field = "genres", sep = ",",
                  similarity = "jaccard", top_n = 10)
genre_top10
```

All three can be combined:

```{r combined-filtering}
co(movies, field = "genres", sep = ",",
   similarity = "association", counting = "fractional",
   min_occur = 15, threshold = 0.001, top_n = 20)
```

## 5. Actor co-occurrence (long/bipartite format)

Actors who share movies are connected. The data is in long format: one row per actor--movie pair. Use `field` for the entity (actor) and `by` for the grouping column (movie ID).

```{r actor-cooccurrence}
actor_net <- co(actor_movie, field = "actor", by = "tconst",
                similarity = "jaccard", min_occur = 3, threshold = 0.1)
actor_net
```

These are actors who frequently appear in each other's "known for" filmographies.

With fractional counting, actors who appear in many movies contribute less per pair:

```{r actor-fractional}
actor_frac <- co(actor_movie, field = "actor", by = "tconst",
                 similarity = "jaccard", counting = "fractional",
                 min_occur = 3, threshold = 0.05)
actor_frac
```

## 6. Splitting by groups

### Genre networks by decade

How do genre pairings change over time? Use `split_by` to compute separate networks per decade:

```{r split-by-decade}
genre_decades <- co(movies, field = "genres", sep = ",",
                    split_by = "decade", similarity = "jaccard",
                    min_occur = 5, top_n = 5)
genre_decades
```

Each decade gets its own top-5 genre pairings. Filter to a specific decade:

```{r filter-decade}
genre_decades[genre_decades$group == "2010s", ]
```

### Actor networks by decade

```{r actor-decades}
actor_decades <- co(actor_movie, field = "actor", by = "tconst",
                    split_by = "decade", similarity = "jaccard",
                    min_occur = 2, top_n = 3)
actor_decades
```

### Split by rating band

Create custom grouping variables for any split:

```{r split-custom}
movies$rating_band <- ifelse(movies$averageRating >= 8, "8+", "7-7.9")
co(movies, field = "genres", sep = ",",
   split_by = "rating_band", similarity = "jaccard",
   min_occur = 10, top_n = 5)
```

## 7. Output formats

### Default (from, to, weight, count)

The standard tidy data frame:

```{r output-default}
co(movies, field = "genres", sep = ",", top_n = 5)
```

### Gephi (Source, Target, Weight, Type)

Ready for Gephi's CSV importer:

```{r output-gephi}
gephi <- co(movies, field = "genres", sep = ",",
            similarity = "jaccard", output = "gephi", top_n = 10)
gephi
```

Export to CSV:

```{r gephi-export, eval=FALSE}
write.csv(gephi, "genre_network_gephi.csv", row.names = FALSE)
```

### igraph

Get an igraph object directly:

```{r output-igraph}
g <- co(movies, field = "genres", sep = ",",
        similarity = "jaccard", min_occur = 20, output = "igraph")
g
```

Full igraph API:

```{r igraph-analysis}
igraph::degree(g)
igraph::betweenness(g)
```

Plot:

```{r igraph-plot, fig.width=8, fig.height=8}
plot(g,
     edge.width = igraph::E(g)$weight * 10,
     vertex.size = igraph::degree(g) * 3,
     vertex.label.cex = 0.8,
     layout = igraph::layout_with_fr(g))
```

### Matrix

Square co-occurrence matrix:

```{r output-matrix}
mat <- co(movies, field = "genres", sep = ",",
          similarity = "jaccard", min_occur = 20, output = "matrix")
round(mat[1:6, 1:6], 3)
```

### cograph

```{r output-cograph, eval=requireNamespace("cograph", quietly=TRUE)}
net <- co(movies, field = "genres", sep = ",",
          similarity = "jaccard", min_occur = 20, output = "cograph")
net
```

## 8. Converters

All converters work on any `cooccurrence` result:

```{r converters}
result <- co(movies, field = "genres", sep = ",",
             similarity = "jaccard", min_occur = 20)

# Square matrix (normalized)
mat <- as_matrix(result)
dim(mat)

# Raw count matrix (before similarity normalization)
raw <- as_matrix(result, type = "raw")
raw[1:5, 1:5]

# igraph
g <- as_igraph(result)
cat("Nodes:", igraph::vcount(g), "| Edges:", igraph::ecount(g), "\n")
```

## 9. Six input formats, one result

The same genre co-occurrence can be computed from four different representations. All produce identical results.

### Delimited field

```{r format-delimited}
res1 <- co(movies, field = "genres", sep = ",")
```

### Long/bipartite table

```{r format-long}
genre_long <- do.call(rbind, lapply(seq_len(nrow(movies)), function(i) {
  gs <- trimws(strsplit(movies$genres[i], ",")[[1]])
  data.frame(movie_id = movies$tconst[i], genre = gs, stringsAsFactors = FALSE)
}))
head(genre_long)

res2 <- co(genre_long, field = "genre", by = "movie_id")
```

### Binary matrix

```{r format-binary}
all_genres <- sort(unique(genre_long$genre))
bin <- matrix(0L, nrow = nrow(movies), ncol = length(all_genres),
              dimnames = list(movies$tconst, all_genres))
for (i in seq_len(nrow(genre_long))) {
  row <- match(genre_long$movie_id[i], movies$tconst)
  bin[row, genre_long$genre[i]] <- 1L
}

res3 <- co(bin)
```

### List of character vectors

```{r format-list}
genre_list <- lapply(strsplit(movies$genres, ","), trimws)
res4 <- co(genre_list)
```

### Verification

```{r format-verify}
cat("Delimited vs Long:  ", all.equal(res1$weight, res2$weight), "\n")
cat("Delimited vs Binary:", all.equal(res1$weight, res3$weight), "\n")
cat("Delimited vs List:  ", all.equal(res1$weight, res4$weight), "\n")
```

## 10. Real-world analysis: What genres define each decade?

Using association strength (which controls for genre frequency) to identify the most distinctive genre pairings per decade:

```{r decade-analysis}
decades_full <- co(movies, field = "genres", sep = ",",
                   split_by = "decade", similarity = "association",
                   min_occur = 5)

# Top 3 genre pairs per decade
top_per_decade <- do.call(rbind, lapply(split(decades_full, decades_full$group), function(d) {
  head(d[order(-d$weight), ], 3)
}))
top_per_decade[, c("group", "from", "to", "weight", "count")]
```

## 11. Complete pipeline

Combining everything --- similarity, fractional counting, scaling, filtering, and direct igraph output --- in one call:

```{r pipeline, fig.width=8, fig.height=8}
g <- co(movies, field = "genres", sep = ",",
        similarity = "jaccard", counting = "fractional",
        scale = "minmax", min_occur = 15, threshold = 0.05,
        output = "igraph")

plot(g,
     edge.width = igraph::E(g)$weight * 8,
     vertex.size = igraph::degree(g) * 2.5,
     vertex.label.cex = 0.8,
     layout = igraph::layout_with_fr(g),
     main = "IMDB Genre Co-occurrence\n(Jaccard, fractional counting, min 15 movies)")
```

