---
title: "Co-occurrence Networks with IMDB Movie Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Co-occurrence Networks with IMDB Movie Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 6,
  message = FALSE, warning = FALSE
)
```

This tutorial demonstrates every feature of the `cooccur` package using 1,000 highly-rated IMDB movies (rating $\geq$ 7.0, $\geq$ 1,000 votes, 1970--2024). We build genre co-occurrence networks, actor collaboration networks, compare decades, apply different similarity measures, and export to Gephi, igraph, and cograph.

## Data

The dataset includes two tables:

- **movies**: 1,000 films with title, year, genres (comma-delimited), decade, rating, and vote count.
- **actor_movie**: ~25,000 actor--movie pairs linking actors to the films they are "known for" on IMDB.

```{r load-data}
library(cooccur)

movies <- read.csv(system.file("extdata/imdb_movies.csv", package = "cooccur"))
actor_movie <- read.csv(system.file("extdata/imdb_actor_movie.csv", package = "cooccur"))

cat(sprintf("Movies: %d | Actor-movie pairs: %d | Unique actors: %d\n",
            nrow(movies), nrow(actor_movie), length(unique(actor_movie$actor))))
head(movies[, c("primaryTitle", "startYear", "genres", "averageRating")])
```

## 1. Genre co-occurrence (delimited field)

The simplest use case. Each movie has a comma-delimited `genres` column. Genres that appear in the same movie are connected.

```{r genre-basic}
genre_net <- cooccurrence(movies, field = "genres", sep = ",")
genre_net
```

Drama dominates because it's the most frequent genre. Let's normalize with Jaccard similarity to control for popularity:

```{r genre-jaccard}
genre_jac <- co(movies, field = "genres", sep = ",", similarity = "jaccard")
genre_jac
```

Now the top edges reflect genuine affinity rather than just frequency. War--History films co-occur more often than chance would predict given their individual frequencies.

### All eight similarity measures compared

```{r similarity-comparison}
sims <- c("none", "jaccard", "cosine", "inclusion",
           "association", "dice", "equivalence")
comparison <- do.call(rbind, lapply(sims, function(s) {
  res <- co(movies, field = "genres", sep = ",", similarity = s, top_n = 3)
  data.frame(similarity = s, from = res$from, to = res$to,
             weight = round(res$weight, 4), count = res$count,
             stringsAsFactors = FALSE)
}))
comparison
```

Notice how `inclusion` gives a weight of 1.0 when one genre almost always appears with another (e.g., Film-Noir is almost always paired with Crime or Drama). Association strength penalizes high-frequency genres the most, surfacing niche pairings.

## 2. Scaling

Raw counts span a wide range. Scaling compresses the distribution for visualization.

```{r scaling}
# Log scaling
genre_log <- co(movies, field = "genres", sep = ",", scale = "log")
head(genre_log)

# Min-max to [0, 1]
genre_mm <- co(movies, field = "genres", sep = ",",
               similarity = "jaccard", scale = "minmax")
head(genre_mm)

# Binary: just presence/absence of co-occurrence
genre_bin <- co(movies, field = "genres", sep = ",", scale = "binary")
head(genre_bin)
```

## 3. Filtering

### Minimum frequency

Drop genres appearing in fewer than 20 movies:

```{r min-occur}
genre_filtered <- co(movies, field = "genres", sep = ",",
                     similarity = "jaccard", min_occur = 20)
genre_filtered
```

### Threshold

Keep only edges with Jaccard similarity above 0.15:

```{r threshold}
genre_strong <- co(movies, field = "genres", sep = ",",
                   similarity = "jaccard", threshold = 0.15)
genre_strong
```

### Top N edges

Keep the 10 strongest connections:

```{r top-n}
genre_top10 <- co(movies, field = "genres", sep = ",",
                  similarity = "jaccard", top_n = 10)
genre_top10
```

## 4. Actor co-occurrence (long/bipartite format)

Actors who share movies are connected. The data is in long format: one row per actor--movie pair.

```{r actor-cooccurrence}
# Actors who share at least 2 movies, Jaccard similarity
actor_net <- co(actor_movie, field = "actor", by = "tconst",
                similarity = "jaccard", min_occur = 3, threshold = 0.1)
actor_net
```

These are actors who frequently appear in each other's "known for" filmographies.

## 5. Splitting by groups

### Genre networks by decade

How do genre pairings change over time? Use `split_by` to compute separate networks per decade:

```{r split-by-decade}
genre_decades <- co(movies, field = "genres", sep = ",",
                    split_by = "decade", similarity = "jaccard",
                    min_occur = 5, top_n = 5)
genre_decades
```

Each decade gets its own top-5 genre pairings. You can filter to a specific decade:

```{r filter-decade}
genre_decades[genre_decades$group == "2010s", ]
```

### Actor networks by decade

```{r actor-decades}
actor_decades <- co(actor_movie, field = "actor", by = "tconst",
                    split_by = "decade", similarity = "jaccard",
                    min_occur = 2, top_n = 3)
actor_decades
```

## 6. Output formats

### Default (from, to, weight, count)

```{r output-default}
co(movies, field = "genres", sep = ",", top_n = 5)
```

### Gephi (Source, Target, Weight, Type)

Ready to paste into Gephi's CSV importer:

```{r output-gephi}
gephi <- co(movies, field = "genres", sep = ",",
            similarity = "jaccard", output = "gephi", top_n = 10)
gephi
```

Export to CSV for Gephi:

```{r gephi-export, eval=FALSE}
write.csv(gephi, "genre_network_gephi.csv", row.names = FALSE)
```

### igraph

```{r output-igraph}
g <- co(movies, field = "genres", sep = ",",
        similarity = "jaccard", min_occur = 20, output = "igraph")
g
```

Full igraph API is available:

```{r igraph-analysis}
igraph::degree(g)
igraph::betweenness(g)
```

Plot with igraph:

```{r igraph-plot, fig.width=8, fig.height=8}
plot(g,
     edge.width = igraph::E(g)$weight * 10,
     vertex.size = igraph::degree(g) * 3,
     vertex.label.cex = 0.8,
     layout = igraph::layout_with_fr(g))
```

### Matrix

```{r output-matrix}
mat <- co(movies, field = "genres", sep = ",",
          similarity = "jaccard", min_occur = 20, output = "matrix")
round(mat[1:6, 1:6], 3)
```

### cograph

```{r output-cograph, eval=requireNamespace("cograph", quietly=TRUE)}
net <- co(movies, field = "genres", sep = ",",
          similarity = "jaccard", min_occur = 20, output = "cograph")
net
```

## 7. Converters on the result

All converters work on any `cooccurrence` result, not just specific output formats:

```{r converters}
result <- co(movies, field = "genres", sep = ",",
             similarity = "jaccard", min_occur = 20)

# Square matrix
mat <- as_matrix(result)
dim(mat)

# Raw count matrix (before similarity normalization)
raw <- as_matrix(result, type = "raw")
raw[1:5, 1:5]

# igraph
g <- as_igraph(result)
cat("Nodes:", igraph::vcount(g), "| Edges:", igraph::ecount(g), "\n")
```

## 8. Multiple input formats with the same data

The same underlying genre co-occurrence can be computed from different representations of the data.

### From the delimited field (default)

```{r format-delimited}
res1 <- co(movies, field = "genres", sep = ",")
```

### From a long/bipartite table

```{r format-long}
# Expand to long format
genre_long <- do.call(rbind, lapply(seq_len(nrow(movies)), function(i) {
  gs <- trimws(strsplit(movies$genres[i], ",")[[1]])
  data.frame(movie_id = movies$tconst[i], genre = gs, stringsAsFactors = FALSE)
}))
head(genre_long)

res2 <- co(genre_long, field = "genre", by = "movie_id")
```

### From a binary matrix

```{r format-binary}
all_genres <- sort(unique(genre_long$genre))
bin <- matrix(0L, nrow = nrow(movies), ncol = length(all_genres),
              dimnames = list(movies$tconst, all_genres))
for (i in seq_len(nrow(genre_long))) {
  row <- match(genre_long$movie_id[i], movies$tconst)
  bin[row, genre_long$genre[i]] <- 1L
}

res3 <- co(bin)
```

### From a list

```{r format-list}
genre_list <- lapply(strsplit(movies$genres, ","), trimws)
res4 <- co(genre_list)
```

All four give identical weights:

```{r format-verify}
cat("Delimited vs Long:  ", all.equal(res1$weight, res2$weight), "\n")
cat("Delimited vs Binary:", all.equal(res1$weight, res3$weight), "\n")
cat("Delimited vs List:  ", all.equal(res1$weight, res4$weight), "\n")
```

## 9. Real-world analysis: What genres define each decade?

```{r decade-analysis, fig.width=10, fig.height=6}
decades_full <- co(movies, field = "genres", sep = ",",
                   split_by = "decade", similarity = "association",
                   min_occur = 5)

# Top 3 genre pairs per decade
top_per_decade <- do.call(rbind, lapply(split(decades_full, decades_full$group), function(d) {
  head(d[order(-d$weight), ], 3)
}))
top_per_decade[, c("group", "from", "to", "weight", "count")]
```

## 10. Putting it all together

A complete analysis pipeline in 5 lines:

```{r pipeline}
co(movies, field = "genres", sep = ",",
   similarity = "jaccard", scale = "minmax",
   min_occur = 15, threshold = 0.1,
   output = "igraph") |>
  plot(edge.width = igraph::E(co(movies, field = "genres", sep = ",",
       similarity = "jaccard", scale = "minmax", min_occur = 15,
       threshold = 0.1, output = "igraph"))$weight * 8,
       vertex.size = 15, vertex.label.cex = 0.7,
       main = "IMDB Genre Co-occurrence (Jaccard, min 15 movies)")
```

## Session info

```{r session}
sessionInfo()
```
