---
title: "Co-occurrence Networks with IMDB Movie Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Co-occurrence Networks with IMDB Movie Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 6,
  message = FALSE, warning = FALSE
)
```

This tutorial demonstrates every feature of the `cooccur` package using 1,000 highly-rated IMDB movies (rating $\geq$ 7.0, $\geq$ 1,000 votes, 1970--2024). We build genre co-occurrence networks, actor collaboration networks, compare decades, apply different similarity measures and counting methods, and export to Gephi, igraph, and cograph.

## Data

```{r load-data}
library(cooccur)
```

## 1. Genre co-occurrence (delimited field)

Each movie has a comma-delimited `genres` column. Genres that appear in the same movie are connected.

```{r genre-basic}
cooccurrence(movies, field = "genres", sep = ",")
```

Drama dominates because it's the most frequent genre. Normalize with Jaccard to control for popularity:

```{r genre-jaccard}
co(movies, field = "genres", sep = ",", similarity = "jaccard")
```

Now the top edges reflect genuine affinity rather than just frequency.

### Comparing similarity measures

Each measure surfaces different structure. Here are the top 3 pairs under each:

```{r sim-none}
co(movies, field = "genres", sep = ",", similarity = "none", top_n = 3)
```

```{r sim-jaccard}
co(movies, field = "genres", sep = ",", similarity = "jaccard", top_n = 3)
```

```{r sim-cosine}
co(movies, field = "genres", sep = ",", similarity = "cosine", top_n = 3)
```

```{r sim-inclusion}
co(movies, field = "genres", sep = ",", similarity = "inclusion", top_n = 3)
```

```{r sim-association}
co(movies, field = "genres", sep = ",", similarity = "association", top_n = 3)
```

```{r sim-dice}
co(movies, field = "genres", sep = ",", similarity = "dice", top_n = 3)
```

```{r sim-equivalence}
co(movies, field = "genres", sep = ",", similarity = "equivalence", top_n = 3)
```

### Which similarity to use?

- **Exploratory work**: Start with `"none"` to see raw counts and understand the data, then try `"jaccard"` or `"cosine"` for a balanced view.
- **Bibliometric and scientometric networks**: `"association"` is recommended by van Eck & Waltman (2009) because it correctly accounts for the expected number of co-occurrences under independence. Two items that are both very frequent will naturally co-occur often; association strength discounts this, revealing which pairs co-occur *more than chance alone would predict*.
- **Detecting hierarchical or subset structure**: `"inclusion"` (the Simpson coefficient) reveals when one item almost always appears with another --- useful for identifying items that are subsets of broader categories, or for detecting dependency relationships.
- **Binary presence/absence networks**: `"jaccard"` or `"dice"` when you only care *whether* items co-occur, not *how often*. Jaccard is stricter (penalizes unbalanced pairs more); Dice is more lenient.
- **Scale-invariant comparison**: `"cosine"` is invariant to absolute frequency --- useful when comparing co-occurrence patterns across datasets of different sizes.
- **Strict filtering**: `"equivalence"` (cosine squared) amplifies differences --- pairs with weak overlap get pushed closer to zero, retaining only the strongest associations.

## 2. Counting methods

By default, each movie contributes equally: if a movie has genres A, B, and C, each pair adds 1. But movies with many genres inflate the network --- a 5-genre movie creates 10 pairs, while a 2-genre movie creates 1.

**Fractional counting** weights each pair by $1/(n-1)$, where $n$ is the number of items.

```{r counting-full}
co(movies, field = "genres", sep = ",", top_n = 5)
```

```{r counting-fractional}
co(movies, field = "genres", sep = ",", counting = "fractional", top_n = 5)
```

Fractional counting is particularly important when some documents have many items while others have few.

## 3. Scaling

Scaling compresses the distribution for visualization or downstream analysis.

```{r scale-log}
co(movies, field = "genres", sep = ",", scale = "log", top_n = 5)
```

```{r scale-minmax}
co(movies, field = "genres", sep = ",", similarity = "jaccard", scale = "minmax", top_n = 5)
```

```{r scale-binary}
co(movies, field = "genres", sep = ",", scale = "binary", top_n = 5)
```

```{r scale-sqrt}
co(movies, field = "genres", sep = ",", scale = "sqrt", top_n = 5)
```

Scaling can be combined with any similarity:

```{r scale-combined}
co(movies, field = "genres", sep = ",",
   similarity = "association", scale = "log", min_occur = 20, top_n = 5)
```

## 4. Filtering

### Minimum frequency

Drop genres appearing in fewer than 20 movies:

```{r min-occur}
co(movies, field = "genres", sep = ",", similarity = "jaccard", min_occur = 20)
```

### Threshold

Keep only edges with Jaccard similarity above 0.15:

```{r threshold}
co(movies, field = "genres", sep = ",", similarity = "jaccard", threshold = 0.15)
```

### Top N

```{r top-n}
co(movies, field = "genres", sep = ",", similarity = "jaccard", top_n = 10)
```

### Combined

```{r combined-filter}
co(movies, field = "genres", sep = ",",
   similarity = "association", counting = "fractional",
   min_occur = 15, threshold = 0.001, top_n = 20)
```

## 5. Actor co-occurrence (long/bipartite format)

Actors who share movies are connected. The data is one row per actor--movie pair. Use `field` for the entity and `by` for the grouping column.

```{r actor-net}
co(actors, field = "actor", by = "tconst",
   similarity = "jaccard", min_occur = 3, threshold = 0.1)
```

With fractional counting:

```{r actor-frac}
co(actors, field = "actor", by = "tconst",
   similarity = "jaccard", counting = "fractional",
   min_occur = 3, threshold = 0.05)
```

## 6. Splitting by groups

### Genre networks by decade

```{r split-decade}
co(movies, field = "genres", sep = ",",
   split_by = "decade", similarity = "jaccard",
   min_occur = 5, top_n = 5)
```

Filter to a specific decade:

```{r filter-decade}
decades <- co(movies, field = "genres", sep = ",",
              split_by = "decade", similarity = "jaccard",
              min_occur = 5, top_n = 5)

decades[decades$group == "2010s", ]

```

### Split by rating band

```{r split-rating}
movies$rating_band <- ifelse(movies$averageRating >= 8, "8+", "7-7.9")
co(movies, field = "genres", sep = ",",
   split_by = "rating_band", similarity = "jaccard",
   min_occur = 10, top_n = 5)
```

## 7. Output formats

### Default

```{r out-default}
co(movies, field = "genres", sep = ",", top_n = 5)
```

### Gephi

```{r out-gephi}
co(movies, field = "genres", sep = ",",
   similarity = "jaccard", output = "gephi", top_n = 10)
```

```{r gephi-export, eval=FALSE}
write.csv(
  co(movies, field = "genres", sep = ",", similarity = "jaccard", output = "gephi"),
  "genre_network.csv", row.names = FALSE
)

```

### cograph (with Gephi layout)

Convert to a cograph network and plot with `splot()` using the Gephi Fruchterman-Reingold layout:

```{r out-cograph, fig.width=8, fig.height=8}
net <- co(movies, field = "genres", sep = ",",
          similarity = "jaccard", min_occur = 20, output = "cograph")
cograph::splot(net, layout = "fr", scale_nodes_by = "degree")

```

With customized edge width and label size:

```{r cograph-styled, fig.width=8, fig.height=8}
cograph::splot(net, layout = "gephi", label_size = .8, label_fontface = "bold",
               node_fill = "#F9C22E",  
               node_border_width = 0.0001, edge_color = "black",
               scale_nodes_by = "degree", edge_width_range = c(0.1:4))
```

### igraph

```{r out-igraph}
g <- co(movies, field = "genres", sep = ",",
        similarity = "jaccard", min_occur = 20, output = "igraph")
g
```

```{r igraph-metrics}
igraph::degree(g)
igraph::betweenness(g)
```

### Matrix

```{r out-matrix}
mat <- co(movies, field = "genres", sep = ",",
          similarity = "jaccard", min_occur = 20, output = "matrix")
round(mat[1:6, 1:6], 3)
```

## 8. Converters

```{r conv-matrix}
result <- co(movies, field = "genres", sep = ",",
             similarity = "jaccard", min_occur = 20)
as_matrix(result)
```

```{r conv-raw}
as_matrix(result, type = "raw")
```

```{r conv-igraph}
as_igraph(result)
```

## 9. Six input formats, one result

The same data in four representations --- all produce identical results.

```{r fmt-delimited}
res1 <- co(movies, field = "genres", sep = ",")
```

```{r fmt-long}
genre_long <- do.call(rbind, lapply(seq_len(nrow(movies)), function(i) {
  gs <- trimws(strsplit(movies$genres[i], ",")[[1]])
  data.frame(movie_id = movies$tconst[i], genre = gs, stringsAsFactors = FALSE)
}))
res2 <- co(genre_long, field = "genre", by = "movie_id")
```

```{r fmt-binary}
all_genres <- sort(unique(genre_long$genre))
bin <- matrix(0L, nrow = nrow(movies), ncol = length(all_genres),
              dimnames = list(movies$tconst, all_genres))
for (i in seq_len(nrow(genre_long))) {
  row <- match(genre_long$movie_id[i], movies$tconst)
  bin[row, genre_long$genre[i]] <- 1L
}
res3 <- co(bin)
```

```{r fmt-list}
res4 <- co(lapply(strsplit(movies$genres, ","), trimws))
```

```{r fmt-verify}
all.equal(res1$weight, res2$weight)
all.equal(res1$weight, res3$weight)
all.equal(res1$weight, res4$weight)
```

## 10. Complete pipeline

Everything in one call --- similarity, fractional counting, scaling, filtering --- piped straight to cograph for visualization:

```{r pipeline, fig.width=8, fig.height=8}
co(movies, field = "genres", sep = ",",
   similarity = "jaccard", counting = "fractional",
   scale = "minmax", min_occur = 15, threshold = 0.05,
   output = "cograph") |>
  cograph::splot(layout = "gephi", edge_width = 3, label_size = 0.9,
                 title = "IMDB Genre Co-occurrence (Jaccard, fractional, min 15 movies)")
```
